### 2017-02-06 introduction.tex
Formatting corrections made

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{irregularity.pdf}
	\caption[Plasma irregularities in the polar ionosphere]{{\:}Plasma irregularities in the polar ionosphere\\ Plasma density irregularities occur in the polar cap at different scale sizes.  Polar patches, orange in Figures \ref{fig:polarirreg}a and \ref{fig:polarirreg}b, move with the background ionospheric convection, indicated by red arrows.  Smaller-scale structures often develop along the edges of polar patches where strong density gradients exist, Figure \ref{fig:polarirreg}b.  Intermediate-scale irregularities are primarily responsible for disrupting radio signals traveling through the ionosphere, Figure \ref{fig:polarirreg}c.  Field-aligned irregularities, FAIs, are density perturbations that form along the magnetic field lines, Figure \ref{fig:polarirreg}d.}
	\label{fig:polarirreg}
\end{figure}

@@

Some of the first 1D investigations of the linear GDI growth rate found that it can be described by the following simple expression involving the plasma drift speed, \(V_E = |\vec{E}\times\vec{B}|/B^2\), and the gradient scale length, \(L^{-1} = |(\partial n/\partial x)/n|\) \citep{Simon1963,Hoh1963,Linson1970}
\begin{equation}
	\label{eqn:gdi_old}
	\gamma = \frac{V_E}{L}.
\end{equation}
This expression is technically only valid if the density gradient, \(\nabla n\), is in the same direction as the plasma drift velocity, \(\vec{V}_E\), and the wave is propagating perpendicular to both.  A more general expression that is used less often does consider an arbitrary wavevector, \(\vec{k}\), but still assumes density gradients parallel to the plasma drift \citep{Tsunoda1988}:
\begin{equation}
	\label{eqn:gdi_80s}
	\gamma = \frac{k_y}{k}\frac{\vec{k}\cdot\vec{V}_E}{kL}.
\end{equation}
An even more general consideration involves arbitrary gradient and plasma drift directions \citep{Keskinen1982a,Makarevich2014c}.  In this more general case, the directional dependence of GDI has been shown to be significant, such that the growth rate can actually 
change signs (determining whether wave growth or damping occurs) with different wavevector and gradient directions.  In addition, these expressions have historically been developed for a particular altitudinal regime, usually in the \(F\) region.

@@

For the most part, the results of nonlinear numerical simulations very closely match expectations from linear GDI theory.  In particular, it was found that the spatial power spectra are anisotropic with most of the power concentrated in the direction where linear GDI growth is most favorable \citep{Keskinen1981a,Keskinen1981b,Keskinen1982a,Gondarenko2001,Gondarenko2004b}.  However, very large shears or stronger ion-inertial factors can cause isotropy \citep{Gondarenko2001,Gondarenko2006}.  The power spectra derived from these simulations are similar to those from experimental observations \citep{Baker1978,Kelley1979}, however it is computationally challenging to run simulations down to decameter scales and observations are generally limited to rocket and satellite techniques, which can only be used to calculate power spectra in one dimension \citep{Villain1986,Moen2012}.

>>>>

\begin{figure}
	\centering
	\includegraphics[height=0.75\textheight]{irregularity.pdf}
	\caption[Plasma irregularities in the polar ionosphere]{{\:}Plasma irregularities in the polar ionosphere\\ Plasma density irregularities occur in the polar cap at different scale sizes.  Polar patches, orange in Figures \ref{fig:polarirreg}a and \ref{fig:polarirreg}b, move with the background ionospheric convection, indicated by red arrows.  Smaller-scale structures often develop along the edges of polar patches where strong density gradients exist, Figure \ref{fig:polarirreg}b.  Intermediate-scale irregularities are primarily responsible for disrupting radio signals traveling through the ionosphere, Figure \ref{fig:polarirreg}c.  Field-aligned irregularities, FAIs, are density perturbations that form along the magnetic field lines, Figure \ref{fig:polarirreg}d.}
	\label{fig:polarirreg}
\end{figure}

@@

Some of the first 1D investigations of the linear GDI growth rate found that it can be described by the following simple expression involving the plasma drift speed, \(V_E = |\vec{E}\times\vec{B}|/B^2\), and the gradient scale length, \(L^{-1} = |(\partial n/\partial x)/n|\) \citep{Simon1963,Hoh1963,Linson1970}
\begin{equation}
	\label{eqn:gdi_old}
	\gamma = \frac{V_E}{L}.
\end{equation}
This expression is technically only valid if the density gradient, \(\nabla n\), is in the same direction as the plasma drift velocity, \(\vec{V}_E\), and the wave is propagating perpendicular to both.  A more general expression that is used less often does consider an arbitrary wavevector, \(\vec{k}\), but still assumes density gradients parallel to the plasma drift \citep{Tsunoda1988}:
\begin{equation}
	\label{eqn:gdi_80s}
	\gamma = \frac{k_y}{k}\frac{\vec{k}\cdot\vec{V}_E}{kL}.
\end{equation}
An even more general consideration involves arbitrary gradient and plasma drift directions \citep{Keskinen1982,Makarevich2014c}.  In this more general case, the directional dependence of GDI has been shown to be significant, such that the growth rate can actually 
change signs (determining whether wave growth or damping occurs) with different wavevector and gradient directions.  In addition, these expressions have historically been developed for a particular altitudinal regime, usually in the \(F\) region.

@@

For the most part, the results of nonlinear numerical simulations very closely match expectations from linear GDI theory.  In particular, it was found that the spatial power spectra are anisotropic with most of the power concentrated in the direction where linear GDI growth is most favorable \citep{Keskinen1981a,Keskinen1981b,Keskinen1982,Gondarenko2001,Gondarenko2004b}.  However, very large shears or stronger ion-inertial factors can cause isotropy \citep{Gondarenko2001,Gondarenko2006}.  The power spectra derived from these simulations are similar to those from experimental observations \citep{Baker1978,Kelley1979}, however it is computationally challenging to run simulations down to decameter scales and observations are generally limited to rocket and satellite techniques, which can only be used to calculate power spectra in one dimension \citep{Villain1986,Moen2012}.

### 2017-02-06 paper1.tex
Formatting corrections made

\chapter[Solar Control of \(F\) Region Radar Backscatter: Further Insights from Observations in the Southern Polar Cap]{Solar Control of \(F\) Region Radar Backscatter: Further Insights from Observations in the Southern Polar Cap\footnote{Published as: {Lamarche}, L.~J., and R.~A. {Makarevich} (2015), Solar control of {F} region radar backscatter: Further insights from observations in the southern polar cap, \textit{J. Geophys. Res. Space Physics}, \textit{120}(11), 9875--9890, \doi{10.1002/2015JA021663}.}}

@@

Production of these irregularities is reasonably well understood on a microphysical level, with GDI promoting irregularity growth for certain orientations of the density gradient vector relative to the convection electric field \citep{Keskinen1982a,Keskinen1982b,Makarevich2014c}. The role of global factors, on the other hand, such as geomagnetic activity or solar wind conditions is much less investigated. In particular, the extent of the solar control is largely unknown, both in terms of its direct effects due to varying ionization and  indirect effects due to varying interplanetary magnetic field (IMF) as explained below. This is particularly important to processes occurring in the polar cap where some of the strongest differences in solar illumination conditions occur and where geomagnetic field lines are directly connected to the IMF.

@@

\subsection{\(\mathbf{F}\) Region Echo Band: A Comprehensive Model and Density Implications}

@@

The backscatter patches move away form MCM in Figure \ref{fig:patchy_example}b.  For example, the patch near 17 UT moved by \(\sim700\) km in \(\sim\)20-min, which translates into \(\sim600\) m/s velocity.  This is consistent with the Doppler velocity measured by MCM within the patch (dark blue colors).  Similar consistency between the patch drift and plasma convection velocities was observed by \citet{Bristow2011} using MCM observations and, more recently, by \citet{Makarevich2015} using direct observations of 2D convection velocity and density enhancements by RISR-N in the northern polar cap.  This impolies that polar patches tend to move with the background convection, at least deep within the polar cap.

@@

\subsection{\(\mathbf{F}\) Region Irregularities: Suppression and Enhancement by Sunlight}

@@

The echo occurrence dependence on IMF in general can be interpreted in context of IMF-controlled convection environment. High-latitude convection patterns are directly controlled by solar wind variations. Statistical convection patterns are derived from satellites and radars and are routinely parameterized by IMF \(B_y\) and \(B_z\) \citep{Haaland2007, Pettigrew2010}. One mechanism through which IMF-controlled variations in convection may be responsible for changes in occurrence of plasma irregularities involves changing orientations between plasma density gradients \(\vec{G}\) and convection velocity \(\vec{V}_E\) vectors. GDI can more easily excite waves when a component \(\vec{G}\cdot\vec{V}_E\) is more positive, with \(\vec{G}\parallel\vec{V}_E\) being most favorable orientation \citep{Keskinen1982a,Makarevich2014c}. IMF-controlled changes in \(\vec{V}_E\) can thus potentially lead to changes in \(\vec{G}\cdot\vec{V}_E\), which, in turn, will change GDI growth rate and occurrence of GDI-produced irregularities. This is a feasible basic scenario, although some questions still remain as to the extent of its contribution to the trends observed and reported here. For example, gradient scales must be relatively small for GDI to contribute to decameter-scale irregularity generation, with one widely-accepted possibility of large-scale gradient structures serving as seeds for small-scale structuring processes \citep{Tsunoda1988}. In this context, the presented results suggest that time scales at which convection changes are shorter than those associated with breakup of large-scale structures and changes in prevalent gradient orientations at both large and small spatial scales. This is consistent with the idea of very fast propagation of convection disturbances \citep{Ridley1998,Ruohoniemi2002,Fiori2012,Taguchi2015}.

>>>>

\chapter[Solar Control of \textit{F} Region Radar Backscatter: Further Insights from Observations in the Southern Polar Cap]{Solar Control of \textit{F} Region Radar Backscatter: Further Insights from Observations in the Southern Polar Cap\footnote{Published as: {Lamarche}, L.~J., and R.~A. {Makarevich} (2015), Solar control of {F} region radar backscatter: Further insights from observations in the southern polar cap, \textit{J. Geophys. Res. Space Physics}, \textit{120}(11), 9875--9890, \doi{10.1002/2015JA021663}.}}

@@

Production of these irregularities is reasonably well understood on a microphysical level, with GDI promoting irregularity growth for certain orientations of the density gradient vector relative to the convection electric field \citep{Keskinen1982,Keskinen1983a,Makarevich2014c}. The role of global factors, on the other hand, such as geomagnetic activity or solar wind conditions is much less investigated. In particular, the extent of the solar control is largely unknown, both in terms of its direct effects due to varying ionization and  indirect effects due to varying interplanetary magnetic field (IMF) as explained below. This is particularly important to processes occurring in the polar cap where some of the strongest differences in solar illumination conditions occur and where geomagnetic field lines are directly connected to the IMF.

@@

\subsection{\textit{F} Region Echo Band: A Comprehensive Model and Density Implications}

@@

The backscatter patches move away form MCM in Figure \ref{fig:patchy_example}b.  For example, the patch near 17 UT moved by \(\sim700\) km in \(\sim\)20-min, which translates into \(\sim600\) m/s velocity.  This is consistent with the Doppler velocity measured by MCM within the patch (dark blue colors).  Similar consistency between the patch drift and plasma convection velocities was observed by \citet{Bristow2011} using MCM observations and, more recently, by \citet{Makarevich2015b} using direct observations of 2D convection velocity and density enhancements by RISR-N in the northern polar cap.  This impolies that polar patches tend to move with the background convection, at least deep within the polar cap.

@@

\subsection{\textit{F} Region Irregularities: Suppression and Enhancement by Sunlight}

@@

The echo occurrence dependence on IMF in general can be interpreted in context of IMF-controlled convection environment. High-latitude convection patterns are directly controlled by solar wind variations. Statistical convection patterns are derived from satellites and radars and are routinely parameterized by IMF \(B_y\) and \(B_z\) \citep{Haaland2007, Pettigrew2010}. One mechanism through which IMF-controlled variations in convection may be responsible for changes in occurrence of plasma irregularities involves changing orientations between plasma density gradients \(\vec{G}\) and convection velocity \(\vec{V}_E\) vectors. GDI can more easily excite waves when a component \(\vec{G}\cdot\vec{V}_E\) is more positive, with \(\vec{G}\parallel\vec{V}_E\) being most favorable orientation \citep{Keskinen1982,Makarevich2014c}. IMF-controlled changes in \(\vec{V}_E\) can thus potentially lead to changes in \(\vec{G}\cdot\vec{V}_E\), which, in turn, will change GDI growth rate and occurrence of GDI-produced irregularities. This is a feasible basic scenario, although some questions still remain as to the extent of its contribution to the trends observed and reported here. For example, gradient scales must be relatively small for GDI to contribute to decameter-scale irregularity generation, with one widely-accepted possibility of large-scale gradient structures serving as seeds for small-scale structuring processes \citep{Tsunoda1988}. In this context, the presented results suggest that time scales at which convection changes are shorter than those associated with breakup of large-scale structures and changes in prevalent gradient orientations at both large and small spatial scales. This is consistent with the idea of very fast propagation of convection disturbances \citep{Ridley1998,Ruohoniemi2002,Fiori2012,Taguchi2015}.

### 2017-02-06 paper2.tex
Formatting corrections made

The GDI growth rate is generally dependent on the relative directions of the background density gradient vector \(\nabla n_0\) and the convection electric field \(\vec{E}\) or, equivalently, the convection velocity \(\vec{V}_E=\vec{E}\times\vec{B}/B^2\). It is widely accepted that the maximum GDI growth rate in the \(F\) region occurs when the density gradient \(\nabla n_0\) is parallel to the convection drift velocity \(\vec{V}_E\) \citep{Keskinen1982a}, while in the \(E\) region the maximum occurs when \(\nabla n_0\parallel \vec{E}\). An analytic expression for the GDI growth rate in the \(F\) region has been derived in numerous studies, with most of these derivations considering a particular subset of mutual orientations between \(\nabla n_0\), \(\vec{V}_E\), and the wave vector \(\vec{k}\). This typically results in the upper limit of the GDI growth rate of \(\gamma=V_E/L\), where \(L\) is a characteristic gradient scale length defined through \(L^{-1}\equiv |\nabla n_0|/n_0\). More general vector configurations and growth rate expressions in the \(F\) region have also been considered \citep{Keskinen1982a,Makarevich2014c}, but their implications for studying the asymmetries between leading and trailing edges of large-scale structures have not been investigated. Further, it has been demonstrated recently that the GDI growth rate patterns change dramatically between the \(E\) and \(F\) regions \citep{Makarevich2014c}. This means that one expects different signatures of small-scale irregularities and their asymmetries near large-scale structures at different altitudinal regions.

@@

Although this study is relevant to observations with obliquely scanning HF radars, care should be taken with how the results are used to interpret experimental data.  In this study, only the linear growth rate of GDI was considered. Even though radar backscatter power and occurrence trends are often interpreted using the GDI growth rates in the linear regime \citep[e.g.][]{Oksavik2012,Moen2012}, there are other factors that may affect echo power and occurrence.  In the polar cap, HF radars rely on beam refraction through a dense ionosphere to observe field-aligned irregularities \citep{Bristow2011,Koustov2014}.  In addition, high background electron density in both the \(E\) and \(F\) regions can smooth gradients required for GDI and limit instability growth \citep{Ruohoniemi1997,Koustov2004}.  Large conductivity in the \(E\) region can also short out \(F\) region irregularities \citep{Danskin2002,Kane2012,Lamarche2015}.  An enhanced D region may act to attenuate backscatter signals, although this effect is not always obvious \citep{Danskin2002}.  Nonlinear processes in the \(F\) region have been theoretically shown to alter the dependence on the propagation direction while still showing significant anisotropy similar to that predicted by the linear theory \citep{Keskinen1981b,Keskinen1982a,Keskinen1983}.  The situation in the \(E\) region appears to be somewhat different, with some echoes also exhibiting strong anisotropy \citep[e.g.][]{Ierkic1980,Makarevich2008} and other echoes showing isotropic power variation \citep{Ierkic1980,Koustov2001} presumably because of nonlinear processes.  The results of the present study are thus more likely to apply in the \(F\) region.

>>>>

The GDI growth rate is generally dependent on the relative directions of the background density gradient vector \(\nabla n_0\) and the convection electric field \(\vec{E}\) or, equivalently, the convection velocity \(\vec{V}_E=\vec{E}\times\vec{B}/B^2\). It is widely accepted that the maximum GDI growth rate in the \(F\) region occurs when the density gradient \(\nabla n_0\) is parallel to the convection drift velocity \(\vec{V}_E\) \citep{Keskinen1982}, while in the \(E\) region the maximum occurs when \(\nabla n_0\parallel \vec{E}\). An analytic expression for the GDI growth rate in the \(F\) region has been derived in numerous studies, with most of these derivations considering a particular subset of mutual orientations between \(\nabla n_0\), \(\vec{V}_E\), and the wave vector \(\vec{k}\). This typically results in the upper limit of the GDI growth rate of \(\gamma=V_E/L\), where \(L\) is a characteristic gradient scale length defined through \(L^{-1}\equiv |\nabla n_0|/n_0\). More general vector configurations and growth rate expressions in the \(F\) region have also been considered \citep{Keskinen1982,Makarevich2014c}, but their implications for studying the asymmetries between leading and trailing edges of large-scale structures have not been investigated. Further, it has been demonstrated recently that the GDI growth rate patterns change dramatically between the \(E\) and \(F\) regions \citep{Makarevich2014c}. This means that one expects different signatures of small-scale irregularities and their asymmetries near large-scale structures at different altitudinal regions.

@@

Although this study is relevant to observations with obliquely scanning HF radars, care should be taken with how the results are used to interpret experimental data.  In this study, only the linear growth rate of GDI was considered. Even though radar backscatter power and occurrence trends are often interpreted using the GDI growth rates in the linear regime \citep[e.g.][]{Oksavik2012,Moen2012}, there are other factors that may affect echo power and occurrence.  In the polar cap, HF radars rely on beam refraction through a dense ionosphere to observe field-aligned irregularities \citep{Bristow2011,Koustov2014}.  In addition, high background electron density in both the \(E\) and \(F\) regions can smooth gradients required for GDI and limit instability growth \citep{Ruohoniemi1997,Koustov2004}.  Large conductivity in the \(E\) region can also short out \(F\) region irregularities \citep{Danskin2002,Kane2012,Lamarche2015}.  An enhanced D region may act to attenuate backscatter signals, although this effect is not always obvious \citep{Danskin2002}.  Nonlinear processes in the \(F\) region have been theoretically shown to alter the dependence on the propagation direction while still showing significant anisotropy similar to that predicted by the linear theory \citep{Keskinen1981b,Keskinen1982,Keskinen1983b}.  The situation in the \(E\) region appears to be somewhat different, with some echoes also exhibiting strong anisotropy \citep[e.g.][]{Ierkic1980,Makarevich2008} and other echoes showing isotropic power variation \citep{Ierkic1980,Koustov2001} presumably because of nonlinear processes.  The results of the present study are thus more likely to apply in the \(F\) region.

### 2017-02-06 paper3.tex
Formatting corrections made

Implicit in the above relationship  between gradient strength vector \(\vec{G} = \nabla n/n\) and the drift velocity \(\vec{V}_E\) (parallel or antiparallel) is an assumption about their typical mutual orientation as well as the associated assumption about directions of the peak growth rate in terms of both the wavevector and the gradient. This leads to significant simplifications in the GDI growth rate expression \citep{Linson1970,Tsunoda1988}. In context of large-scale density structures at high-latitudes drifting perpendicular to their direction of elongation \citep{Makarevitch2004c,Makarevich2015b} such as polar cap patches and sun-aligned arcs, \(\vec{G}\) and \(\vec{V}_E\) are parallel on the trailing edge and antiparallel on the leading edge. However, there can be large variations in this \citep{Makarevitch2004c} and, in general, the growth rate changes with the wavevector direction \citep{Keskinen1982a,Keskinen1983,Makarevich2014c}. More insight can therefore be gained by considering the directions of vector quantities instead of considering a vector configuration that results in the peak growth.

@@

\subsection{Effects of Density-Related Factors}
\label{sec:p3discussion1}
The fluid linear theory of GDI predicts that an increase in plasma density gradient strength \(G\) will increase irregularity production through GDI \citep{Keskinen1983,Makarevich2014c}. The current experimental results were generally consistent with this expectation. In particular, the echo power showed an expected increase with \(G\), Figure \ref{fig:density}c. The point scatter was large but the increase was monotonic for bins with significant statistics. One of  most interesting aspects of these observations arguably concerns small scales of irregularities considered (\(\lambda \approx 10\) m), as explained below.

@@

\subsection{Directional Effects of Electric Field and Gradient Components}
\label{sec:p3discussion2}
One important new aspect of the current study is the experimental investigation of the directional dependence of the GDI growth rate, as represented by equation  (\ref{eqn:p3gdi}). This directional dependence has been previously predicted theoretically \citep{Keskinen1982a,Makarevich2014c,Lamarche2016}, but no experimental evidence is currently available, apart from asymmetry between leading and trailing edges of large-scale density structures. Because the growth rate \(\gamma\) is directly dependent on both the gradient component, \(\uvec{k}\cdot\uvec{b}\times\vec{G}\), and the electric field component, \(\uvec{k}\cdot\vec{E}\), the relative directions of the gradient, electric field, and wavevector are predicted to strongly influence plasma structuring in the ionosphere, equation (\ref{eqn:p3gdi}).  A new method was used to find 2D plasma density gradients based on RISR-N measurements and these, along with electric field measurements, are used here to examine directional dependencies.

>>>>

Implicit in the above relationship  between gradient strength vector \(\vec{G} = \nabla n/n\) and the drift velocity \(\vec{V}_E\) (parallel or antiparallel) is an assumption about their typical mutual orientation as well as the associated assumption about directions of the peak growth rate in terms of both the wavevector and the gradient. This leads to significant simplifications in the GDI growth rate expression \citep{Linson1970,Tsunoda1988}. In context of large-scale density structures at high-latitudes drifting perpendicular to their direction of elongation \citep{Makarevitch2004c,Makarevich2015b} such as polar cap patches and sun-aligned arcs, \(\vec{G}\) and \(\vec{V}_E\) are parallel on the trailing edge and antiparallel on the leading edge. However, there can be large variations in this \citep{Makarevitch2004c} and, in general, the growth rate changes with the wavevector direction \citep{Keskinen1982,Keskinen1983b,Makarevich2014c}. More insight can therefore be gained by considering the directions of vector quantities instead of considering a vector configuration that results in the peak growth.

@@

\subsection{Effects of Density-Related Factors}
\label{sec:p3discussion1}
The fluid linear theory of GDI predicts that an increase in plasma density gradient strength \(G\) will increase irregularity production through GDI \citep{Keskinen1983b,Makarevich2014c}. The current experimental results were generally consistent with this expectation. In particular, the echo power showed an expected increase with \(G\), Figure \ref{fig:density}c. The point scatter was large but the increase was monotonic for bins with significant statistics. One of  most interesting aspects of these observations arguably concerns small scales of irregularities considered (\(\lambda \approx 10\) m), as explained below.

@@

\subsection{Directional Effects of Electric Field and Gradient Components}
\label{sec:p3discussion2}
One important new aspect of the current study is the experimental investigation of the directional dependence of the GDI growth rate, as represented by equation  (\ref{eqn:p3gdi}). This directional dependence has been previously predicted theoretically \citep{Keskinen1982,Makarevich2014c,Lamarche2016}, but no experimental evidence is currently available, apart from asymmetry between leading and trailing edges of large-scale density structures. Because the growth rate \(\gamma\) is directly dependent on both the gradient component, \(\uvec{k}\cdot\uvec{b}\times\vec{G}\), and the electric field component, \(\uvec{k}\cdot\vec{E}\), the relative directions of the gradient, electric field, and wavevector are predicted to strongly influence plasma structuring in the ionosphere, equation (\ref{eqn:p3gdi}).  A new method was used to find 2D plasma density gradients based on RISR-N measurements and these, along with electric field measurements, are used here to examine directional dependencies.
